Sequential algorithms to split and merge terabyte-scale 3D images

### 2017-06-03 paper.tex
Added a README
Updates

@@

\title{Sequential algorithms to split and merge large 3D images}

@@

\begin{abstract}
  \todo{The abstract goes here. DO NOT USE SPECIAL CHARACTERS,
    SYMBOLS, OR MATH IN YOUR TITLE OR ABSTRACT.}  Authors (final order
  to be discussed when the paper shapes up): Yongping/Val\'erie,
  Yuhong Yan, Tristan. Perhaps Pierre Bellec.
\end{abstract}

@@

Three-dimensional images that cannot be loaded in memory are
increasingly found in a variety of disciplines. BigBrain, for
instance, is a 3D histological image of the human brain that
represents 1~TB of data organized in 3600 planes at full resolution
(1$\mu$m in-plane), and 76~GB at a 40$\mu$m isotropic resolution
commonly used in neurosciences~\cite{amunts2013bigbrain}. Other
examples found in medical imaging, our primary domain of interest,
include \todo{electromicroscopy,spectroscopy-francoise-peyrin}. As
such images would typically be processed on a computing cluster,
possibly using locality-aware file systems such as HDFS, software
libraries are needed to split and merged such images efficiently. We
introduce and compare a family of algorithms to address this problem
in a sequential environment made of a single storage disk. We are
aware of the need for parallel algorithms that would split and merge
images to and from an array of disks but we plan to describe such
algorithms in a subsequent paper based on the results of the present
one.

A disk is modeled by its read and write rates ($\alpha_r$ and
$\alpha_w$, in MB/s), and its seek speed ($\beta$, also in
MB/s).

@@

\section{Algorithms and Models}

In our context, split and merge relate to the same dual
problem \todo{not so sure about this actually}. Without loss of
generality, we focus here on merging for the sake of concision. Our
goal then is to merge a set of $N$ chunks into a single reconstructed
3D image with $F$ voxels of size $u$. We assume that all chunks are of
identical size and that slices are squares and blocks are cubes.

Without loss of generality, we consider a file format where voxels are
written after alphanumerical sorting of their (x,y,z)
representation. That is, x slices are written one after the other, and
in every x slice, y rows are written one after the other. NifTI, the
file format defined by the Neuroimaging Informatics Technology
Initiative Nifti\footnote{\url{https://nifti.nimh.nih.gov}} is such a
format, used by the vast majority of the neuroimaging community. In
the remainder, we use the term "slice" for "x slice".

@@

We adopt the following notations:
\begin{itemize}
\item $R$: number of voxels in the reconstructed images.
\item $b$: bytes per voxel in the reconstructed images and splits (in B).
\item $n$: number of splits (blocks or slices).
\item $\alpha_r$: data read rate from disk (in B/s).
\item $\alpha_w$: data write rate to disk (in B/s).
\item $\beta$: disk access time (in s).
\item $m$: amount of RAM available (in B).
\end{itemize}

The following properties are straightforward applications of these
definitions:
\begin{itemize}
\item bytes per chunk: $\frac{Rb}{n}$
\item number of chunks that can fit in memory: $\frac{m.n}{Rb}$
\end{itemize}
For the sake of mathematical simplicity, the models below assume that
the reconstructed image and blocks are cubic. Under this assumption,
we have the following definitions and properties:
\begin{itemize}
\item Number of slices in a block: $\sqrt[3]{\frac{R}{n}}$
\item Number of blocks required to fully cover a slice in the
  reconstructed image: $\sqrt[3]{n}^2=n^{\frac{2}{3}}$
  \item Number of full slices that can be reconstructed from $k$
    blocks: $\floor*{\frac{k}{n^{\frac{2}{3}}}}\sqrt[3]{\frac{R}{n}}$
\end{itemize}
\todo{Make a figure to illustrate the notations.}

@@

Algorithms~\ref{algo:naive-slices} and~\ref{algo:naive-blocks} are the
naive solutions to reconstruct large images from slices or blocks.
\begin{algorithm}{tbh}
\caption{Naive merging of slices.}
\label{algo:naive-slices} 
\begin{algorithmic}
  \FOR{each slice}
    \STATE read slice
    \STATE write slice in reconstructed image
  \ENDFOR      
\end{algorithmic}
\end{algorithm}

\begin{algorithm}{tbh}
\caption{Naive merging of blocks.}
\label{algo:naive-blocks}
\begin{algorithmic}
  \FOR{each block}
    \STATE read block
    \STATE write block in reconstructed image
  \ENDFOR 
\end{algorithmic}
\end{algorithm}

Although simplistic, Algorithms~\ref{algo:naive-slices}
and~\ref{algo:naive-blocks} actually exhibit very different
performance due to the amount of seeks performed by Algorithm 1.b. The
number of seeks performed in Algo 1.b for each block is indeed the
number of rows in a slice, that is, $(R/n)^{1/3}$. The total number of
seeks, considering $n$ blocks is thus $N^{2/3}.F^{1/3}$. In practice,
this could lead to a tremendous slowdown.

For slices, the total merge time is expressed as follows:
\begin{equation}
T_\mathrm{slices} = \frac{bR}{\alpha_r}+\frac{bR}{\alpha_w}+2\beta n
\end{equation}
and for blocks:
\begin{equation}
  T_\mathrm{blocks} = \frac{bR}{\alpha_r}+\frac{bR}{\alpha_w}+\beta n\left(1+\sqrt[3]{\frac{R}{n}}\right)
\end{equation}

@@

\section{Results}

@@

\section{Discussion}

@@

File formats:
\begin{itemize}
\item would MINC~\cite{vincent2016minc} support such algorithms? See on-going email discussion
with P. Bellec.
\item some formats optimize the storage for particular split shapes. See
  ndstore~\cite{burns2013open}. Such formats would behave equally
  poorly with splits that do not comply to this geometry. Our
  algorithms would work equally well with such formats.
\end{itemize}

Reconstructing (therefore splitting) a complete 3D image may not
always be necessary. Re-splitting, for instance, could be done from
existing splits.

>>>>

\newcommand{\note}[1]{
  \color{blue}\emph{[Note: #1]}
  \color{black}
}

@@

\title{Sequential algorithms to split and merge terabyte-scale 3D images}

@@

\begin{abstract}
  \todo{The abstract goes here. DO NOT USE SPECIAL CHARACTERS,
    SYMBOLS, OR MATH IN YOUR TITLE OR ABSTRACT.}  Authors (final order
  to be discussed when the paper shapes up): Yongping/Val\'erie,
  Yuhong Yan, Tristan. Perhaps Pierre Bellec. \note{This is a note.}
\end{abstract}

@@

Three-dimensional images that exceed typical memory size are
increasingly found in a variety of disciplines. BigBrain, for
instance, is a 3D histological image of the human brain that
represents 1~TB of data organized in 3600 planes at full resolution
(1$\mu$m in-plane), and 76~GB at a 40$\mu$m isotropic resolution
commonly used in neurosciences~\cite{amunts2013bigbrain}. Other
examples found in medical imaging, our primary domain of interest,
include \todo{electromicroscopy,spectroscopy-francoise-peyrin}. As
such images would typically be processed on a computing cluster,
possibly using locality-aware file systems such as HDFS, software
libraries are needed to split and merged such images efficiently. We
introduce and compare a family of algorithms to address this problem
in a sequential environment made of a single storage disk. We are
aware of the need for parallel algorithms that would split and merge
images to and from an array of disks but we plan to describe such
algorithms in a subsequent paper based on the results of the present
one.

A disk is modeled by its read and write rates ($\alpha_r$ and
$\alpha_w$, in MB/s), and its access time ($\beta$, in seconds).
%also in
%MB/s).

@@

This paper makes the following contributions:
\begin{itemize}
  \item We propose a set of algorithms to split and merge 3D images
    that do not fit in memory.
  \item We model the performance (execution time) of these algorithms,
    based on a simple but realistic disk I/O model.
  \item We evaluate our model and algorithms on a representative
    use-case involving a 1~TB image a hard drive and a solid-state
    drive.
  \item We compare the algorithms in a variety of conditions based on the validated model.
\end{itemize}
Section~\ref{sec:algos} presents our algorithms and models,
Section~\ref{sec:results} details the results and
Sections~\ref{sec:discussion} discusses them.

\section{Algorithms and Models}
\label{sec:algos}

In our context, split and merge relate to the same dual
problem \todo{not so sure about this actually}. Without loss of
generality, we focus here on merging for the sake of concision. Our
goal then is to merge a set of $n$ chunks into a single reconstructed
3D image with $R$ voxels of size $b$. We assume that all chunks are of
identical size and that slices are squares and blocks are cubes.

Without loss of generality, we consider a file format where voxels are
written after alphanumerical sorting of their (x,y,z)
representation. That is, x slices are written one after the other, and
in every x slice, y rows are written one after the other. NifTI, the
file format defined by the Neuroimaging Informatics Technology
Initiative Nifti\footnote{\url{https://nifti.nimh.nih.gov}} is such a
format, used by the vast majority of the neuroimaging community. In
the remainder, we use the term ``slice" for ``x slice".

@@

We adopt the following notations:
\begin{itemize}
\item $R$: number of voxels in the reconstructed image.
\item $b$: number of bytes per voxel in the reconstructed image and chunks (in B).
\item $n$: number of chunks (blocks or slices).
\item $\alpha_r$: data read rate from disk (in B/s).
\item $\alpha_w$: data write rate to disk (in B/s).
\item $\beta$: disk access time (in s).
\item $m$: amount of available RAM (in B).
\end{itemize}

The following properties are straightforward applications of these
definitions:
\begin{itemize}
\item number of bytes per chunk: $\frac{Rb}{n}$
\item number of chunks that can fit in memory: $\frac{m.n}{Rb}$
\end{itemize}
For the sake of mathematical simplicity, the models below assume that
the reconstructed image and blocks are cubic. Under this assumption,
we have the following definitions and properties:
\begin{itemize}
\item Number of slices in a block: $\sqrt[3]{\frac{R}{n}}$
\item Number of blocks required to fully cover a slice in the
  reconstructed image: $\sqrt[3]{n}^2=n^{\frac{2}{3}}$
  \item Number of full slices that can be reconstructed from $k$
    blocks: $\floor*{\frac{k}{n^{\frac{2}{3}}}}\sqrt[3]{\frac{R}{n}}$
\end{itemize}
\todo{Make a figure to illustrate the notations.}

@@

Algorithms~\ref{algo:naive-slices} and~\ref{algo:naive-blocks} are the
naive merging methods for slices and blocks.
\begin{algorithm}[h]
\caption{Naive merging of slices.}
\label{algo:naive-slices} 
\begin{algorithmic}
  \FOR{each slice}
    \STATE read slice
    \STATE write slice in reconstructed image
  \ENDFOR      
\end{algorithmic}
\end{algorithm}

\begin{algorithm}[h]
\caption{Naive merging of blocks.}
\label{algo:naive-blocks}
\begin{algorithmic}
  \FOR{each block}
    \STATE read block
    \STATE write block in reconstructed image
  \ENDFOR 
\end{algorithmic}
\end{algorithm}

Algorithms~\ref{algo:naive-slices} and~\ref{algo:naive-blocks}
actually have very different performance even though blocks and slices
have identical sizes. This is due to the number of seeks performed by
Algorithm~\ref{algo:naive-blocks} to write a block, which equals the
total number of rows in all slices of a block, that is (since the
blocks are cubic): $\sqrt[3]{\frac{R}{n}}^2$. \note{Note the square
  which is omitted in equation 2 below and in model.py. Adding the
  square to the model just makes the runtime aberrant. This is because
  not all seeks are equal: only ``long'' seeks between block slices
  are significant; seeks between rows of a given block slice are much
  shorter. The way to model this is to replace $\beta$ by $\beta +
  \frac{s}{\gamma}$ where $\beta$ is in seconds, $\gamma$ is in B/s
  and $s$ is the seek ``distance'' in bytes. I bet $\gamma$ is very
  high on the SSD. } In practice, this could lead to a tremendous
slowdown as we will show later. For slices, the total merge time is
modeled as follows:
\begin{equation}
T_\mathrm{slices} = \frac{bR}{\alpha_r}+\frac{bR}{\alpha_w}+2\beta n
\end{equation}
and for blocks:
\begin{equation}
  T_\mathrm{blocks} = \frac{bR}{\alpha_r}+\frac{bR}{\alpha_w}+\beta n\left(1+\sqrt[3]{\frac{R}{n}}\right)
\end{equation}

@@

\section{Results}
\label{sec:results}

@@

\section{Discussion}
\label{sec:discussion}

@@

File formats:
\begin{itemize}
\item would MINC~\cite{vincent2016minc} support such algorithms? See on-going email discussion
with P. Bellec.
\item some formats optimize the storage for particular split shapes. See
  ndstore~\cite{burns2013open}. Such formats would behave equally
  poorly with chunks that do not comply to this geometry. Our
  algorithms would work equally well with such formats.
\end{itemize}

The performance model might be dependent on the particular
infrastructure (disks, OS and filesystem) used in the experiments. In
particular, caching might behave differently in a different setup.

Reconstructing (therefore splitting) a complete 3D image may not
always be necessary. Re-splitting, for instance, could be done from
existing chunks.
